Lineage tracing and live-cell imaging reveal that NeuroD1 does not reprogram microglia into neurons

Version 1

Abstract

Induction of glia-to-neuron conversion is a promising regenerative strategy for treating brain injuries and neurodegenerative diseases. Previous studies have suggested that NeuroD1 can induce microglia-to-neuron cross-lineage conversion. However, it remains highly controversial. To conclusively determine whether NeuroD1 can convert microglia into neurons, we used genetic fate mapping to track the cell fate of microglia ectopically expressing NeuroD1. Furthermore, we performed two-photon imaging to trace the fate of the NeuroD1-expressing microglia. Our findings revealed that cells ectopically expressing NeuroD1 are bona fide microglia, not neurons, regardless of injury preconditioning. Additionally, NeuroD1 overexpression in microglia did not promote brain injury recovery, and the microglial identity remained intact. These results provide solid evidence that NeuroD1 cannot convert microglia into neurons. Our study underscores the necessity of lineage-tracing and cell fate mapping strategies to verify glia-to-neuron conversion, highlighting the importance of rigorous validation in regenerative research.

Introduction

Neurons in the central nervous system (CNS) possess limited regenerative capacity, resulting in permanent neuronal loss after injury or damage. In contrast, glia especially microglia exhibit a higher regenerative capacity(1–3). Reprogramming glial cells with regenerative potential into neurons lacking regenerative capacity offers a promising strategy for repairing damage in the CNS. The glia-to-neuron conversion process is achieved primarily by the overexpression of specific fate-determining factors or the knockdown of certain genes. The manipulation of single genes such as NeuroD1 (4), PAX6 (5), SOX2 (6–9), ASCL1 (10) and PTBP1 (11, 12) has been reported to induce inter-neuroectodermal lineage conversion, transforming astrocytes, oligodendrocyte precursor cells (OPCs) or Muller glia into neurons. While this approach may hold significant potential for clinical therapies, it still requires validation through more stringent and robust approaches. Well-controlled lineage tracing and unambiguous live imaging are gold standards for demonstrating the glia-to-neuron conversion (13, 14). However, controversial studies employing rigorous lineage tracing and live imaging techniques have failed to observe some inter-lineage glia-to-neuron conversions under single-gene manipulation (15–23), including NeuroD1 (21–23). In addition to inter-lineage conversion, Nakashima et al. have reported a NeuroD1-induced microglia-to-neuron cross-lineage conversion (24). Using lentiviral overexpression of NeuroD1, we conducted lineage tracing, in vitro live-cell imaging, and microglia ablation, demonstrating that the previously reported cross-lineage conversion of microglia into neurons is a result of experimental artifacts, e.g., off-target transduction from virus (13). Although Nakashima et al. argued that the phenomenon we observed was due to insufficient NeuroD1 expression in microglia (25), our comparison of the data from both studies (13, 24) indicates that our NeuroD1 expression levels are comparable to, or even higher than, those reported in their study (26).

Surprisingly, Nakashima et al. recently reported that NeuroD1 can not only can achieve the cross-lineage conversion of microglia into neurons at brain injury sites in vivo, but also promotes recovery from brain injury (27). The implementation of novel yet contentious strategies in clinical trials or practice requires extreme caution. Such conflicting reports are of particular concern because they directly influence the potential translation of glia-to-neuron reprogramming strategies into therapeutic applications. To address this controversy, we have employed a comprehensive combination of methodologies, including genetic lineage tracing (virus-independent to prevent potential leakage) and two-photon in vivo live imaging (enabling direct observation of NeuroD1-expressing microglial cell fate). Our findings unequivocally demonstrate that microglia do not undergo conversion into neurons in vivo under either physiological or pathological conditions. These results underscore the critical importance of employing rigorous lineage tracing and cell fate mapping techniques to validate glial reprogramming approaches.

Results

Microglia-specific NeuroD1 expression using a virus-free genetic lineage tracing system

In our previous study, lineage tracing revealed that lentivirus-driven ectopic expression of NeuroD1 does not induce the conversion of microglia into neurons but rather triggers microglial cell death through BCL2-dependent apoptosis (28). However, the viral tools used demonstrated limited efficiency in specifically targeting microglia in vivo and resulted in off-target neuronal expression, regardless of whether lentivirus or adeno-associated virus (AAV) (with different serotypes 6, 8 and 9) was employed, and irrespective of the use of CX3CR1 or CD68 promoters (Figure S1). Similar observations have also been reported by our prior study and those of other groups (21, 28, 29).

Direct evidence for glia-to-neuron conversion by NeuroD1 via a virus-free ectopic expression system genetic lineage tracing is lacking, by which avoids experimental artifacts from virus leakage. Thus, we utilized transgenic animals to ectopically express NeuroD1 in microglia and track the cell fate of microglia expressing NeuroD1. We crossed a microglia-specific mouse line CX3CR1-CreER (30) with an inducible NeuroD1-expressing mouse line Rosa26-LoxP-Stop-LoxP-NeuroD1-IRES-GFP (LSL-NeuroD1-GFP) (31), to obtain CX3CR1-CreER::LSL-NeuroD1-GFP mice(NeuroD1MG for short). After tamoxifen treatment, NeuroD1 was ectopically expressed in microglia, accompanied by GFP for lineage tracing (Figure 1A). After tamoxifen administration, the majority of GFP+ cells faithfully express NeuroD1 in both the cortex and the striatum across time (Figure 1B-D, Figure S2). Results showed that GFP was specifically expressed in microglia avoiding off-target expression in neurons (Figure 1E-H).

Genetic gain of function of NeuroD1 does not result in microglia-to-neuron conversion in the adult mouse brain.

(A) Schematic of the specific expression of NeuroD1 and lineage tracing of microglia. (B) Experimental timeline. (C) Representative confocal images at D18 showing that NeuroD1 is robustly co-expressed in GFP-positive cells after Cre recombination. (D) Quantification of the percentage of NeuroD1+GFP+ cells among GFP+ cells and NeuroD1+GFP- cells among NeuroD1+ at D18, revealing NeuroD1 is highly co-expressed with GFP. N = 5 mice for each group. (E) Representative confocal images at D18 showing that GFP-positive cells do not colocalize with neuronal marker NeuN. (F) Quantification of NEUN+GFP+ double-positive cells in NEUN+ cells at D18. N = 6 mice for each group. (G) Representative confocal images at D18 showing that GFP-positive cells do colocalize with microglial marker IBA1. (H) Quantification of IBA1+GFP+ double-positive cells in GFP+ cells at D18. N = 5 mice for each group. Data are presented as mean ± SD.

NeuroD1 fails to induce microglia-to-neuron conversion in vivo

Based on this stringent lineage tracing tool, all GFP⁺ cells remained IBA1⁺ microglia across all examined time points (4, 10, 18, 30, and 45 days post–tamoxifen induction; Figure 1E, F, Figure 2B,D). Consistently, no GFP⁺NeuN⁺ double-positive cells were detected, indicating that ectopic expression of NeuroD1 did not induce microglia-to-neuron conversion (Figure 1G,H, Figure 2B,D). Besides, GFP+ cells were also not co-labeled with MAP2 or TUJ1(Figure S3). Therefore, the genetic lineage tracing experiment reveals that NeuroD1 cannot drive microglia-to-neuron conversion in vivo.

Ectopic NeuroD1 expression in microglia does not induce microglia-to-neuron conversion across time.

(A) Experimental timeline. (B) Representative confocal images at D4, D10, D30 and D45 showing that GFP is co-labeled with IBA1. (C) Representative confocal images at D4, D10, D30 and D45 showing that GFP is not co-labeled with NeuN. (D) Quantification of IBA1+GFP+ double-positive cells among GFP+ cells at D4, D10, D30 and D45. N = 4 or 5 mice for each group. (D) Quantification of NeuN+GFP+ double-positive cells among GFP+ cells at D4, D10, D30 and D45. N = 4 or 5 mice for each group.

NeuroD1 expression induces microglial loss but not conversion in vivo

Since immunostaining might yield false-positive signals in the absence of true expression, we further examined microglial states in vivo using 2-photon microscopy to track NeuroD1-expressing, GFP-labeled microglia. Imaging of GFP+ cells in the cortex of NeuroD1MG mice was performed every 4 days starting from 18 days post–tamoxifen induction (Figure 3A). Through 2-photon imaging, we did not observe any morphological evidence of microglia-to-neuron conversion. Instead, the GFP-expressing cells exhibited highly ramified morphologies with long processes characteristic of microglia, rather than typical morphology of neuronal cells (Figure 3B, Movie 1). Notably, some GFP-labeled microglia disappeared over time(Figure 3B, yellow arrow), suggesting that NeuroD1 expression does not induce microglia to neuron conversion but instead promotes microglial cell death in vivo, consistent with our previous in vitro findings(32).To further assess this effect, we quantified GFP+ microglia over time after tamoxifen induction. The number of GFP+ cells initially increased and peaked at D30, but subsequently declined by D45 (Figure 3C, D), supporting the conclusion that NeuroD1 expression contribute to microglia loss in vivo. To further investigate the mechanisms underlying microglial loss following NeuroD1 overexpression, we examined markers of autophagy in cortical sections at D18. In NeuroD1MG mice, reporter+ cells exhibited about 4-fold higher LC3 and P62 levels than controls, as shown by immunostaining (Figure 3E-G), indicating accumulation of autophagosomal markers and suggesting impaired autophagic flux. Consistent with increased cell death, TUNEL staining revealed the increase of TUNEL+ cells among reporter+ cells in NeuroD1MG mice relative to controls (Figure S4). Taken together, these data indicate that NeuroD1 overexpression triggers cellular stress and death pathways (including autophagy dysfunction and apoptosis), leading to microglial loss in vivo rather than microglia-to-neuron conversion.

NeuroD1-expressing microglia retain microglial morphology and decline in number over time.

(A) Experimental timeline. (B) Representative images showing GFP-positive cells with highly ramified morphologies characteristic of microglia rather than neurons. (C) Represent images illutrating changes in the number of GFP-postive cells over time (green dots indicate GFP-positive cell). (D) Quantification of GFP-positive cells density.N = 4 - 6 mice per time point. Data are presented as mean ± SD. One Way ANOVA with Turkey’s multiple comparison test. (E) Represent confocal image shows the LC3 and P62 level in control and NeuroD1MG mice. Scale bar = 50μm. (F) Quantification of P62+ area per cell. Unpaired t test. (G) Quantification of LC3+ area per cell. Unpaired t test. (H) Data are presented as mean ± SD.

NeuroD1 fails to reprogram microglia into neurons even after traumatic brain injury

Previous studies have suggested that injury-induced environments might facilitate glial cell fate reprogramming. For example, Heinrich et al. demonstrated that a stab wound prior to virus delivery is required for SOX2-mediated conversion of NG2 glia(33). Irie et al. showed that NeuroD1 can directly reprogram microglia/macrophages into neurons at the lesion site following transient middle cerebral artery occlusion(27). Likewise, Chen et al. reported that NeuroD1 converts astrocytes into neuron after spinal cord injury (SCI). These findings suggest that the failure of NeuroD1-mediated microglia-to-neuron conversion under physiological conditions may result from an unfavorable microenvironment.

To test the possibility, we subjected NeuroD1MG mice to traumatic brain injury (TBI) in the motor cortex, followed by tamoxifen induction, and utilized Cx3cr1-creER::Ai14 (Ai14MG) mice as controls (Figure 4A). Firstly, we performed grip strength task and rotarod tests to evaluate whether NeuroD1 expression in microglia could ameliorate motor deficits in TBI (Figure 4A). In grip strength task, the NeuroD1-overexpression group showed comparable performance to controls, with no significant differences in the strength across groups (Figure 4B). Similarly, no improvements were observed in rotarod task, reflected by the latency to fall (Figure 4C), indicating that NeuroD1 expression in microglia does not rescue TBI-induced motor dysfunction.

NeuroD1 does not induce microglia-to-neuron conversion even in TBI model.

(A) Experiment design. (B) Quantification of grip strength showing no improvement in NeuroD1-expressing TBI mice. Two-way ANOVA with multiple comparisons. (C) Quantification of rotarod performance showing no improvement in NeuroD1-expressing mice. Two-way ANOVA with multiple comparisons. (D) Representative confocal images of DAPI, GFP and NeuN co-staining results in NeuroD1MG TBI mice, showing no co-localization of GFP with NeuN in either the lesion core or distal regions. Scale bar, 50 μm. (E) Quantification of the percentage of GFP+NeuN+ cells among total NeuN+ cells in both the lesion area and distal area. n.d., not detected. (F) Representative confocal images of DAPI, GFP, and MAP2 in NeuroD1MG mice, showing no co-localization of GFP with MAP2 in either the lesion core or distal regions. Scale bar, 50 μm. (G) Quantification of the percentage of GFP+MAP2+ cells among total MAP2+ cells in the lesion area and distal region. n.d., not detected. (H) UMAP reduction shows the cell types of the FACS-sorted reporter+ cells in each group; note that there are no neurons. (I) Feature plot showing that the microglia-specific marker P2ry12 was highly expressed even under TBI conditions, but the neuron-specific marker Rbfox3 was barely expressed. (J) Violin plot showing the expression of microglia-specific genes (Hexb and Tmem119) and neuron-specific genes (Tubb3 and Map2) in the collected cells.

To further investigate whether NeuroD1 expression could induce microglia-to-neuron conversion, we performed immunostaining on these brains. In both lesion core and distal regions, the GFP-labeled NeuroD1-expressing microglia did not co-localize with the neuronal marker NeuN (Figure 4D, E) and MAP2 (Figure 4F, G), similar to what was observed in control group (data not shown). Besides, we collected reporter (tdTomato or GFP) positive cells in both physiological and TBI condition for scRNA-seq, On the basis of cell type-specific annotations, we identified the cell types of the sorted cells and found no evidence of neuronal clusters in the scRNA-seq results (Figure 4H). Additionally, the microglial markers P2ry12, Hexb, and Tmem119 remained highly expressed in reporter+ microglia (Figure 4I, J), whereas neuronal markers such as Rbfox3, Tubb3, and Map2 were absent in both control and NeuroD1-overexpressing microglia (Figure 4I,J). Overall, these findings demonstrate that NeuroD1 fails to induce microglia-to-neuron conversion, even in the context of TBI.

Discussion

Our study provides compelling evidence that ectopic expression of NeuroD1 does not induce microglia-to-neuron conversion in the adult mouse brain. By employing virus-independent genetic lineage tracing and two-photon in vivo imaging, we systematically assessed the cell fate of NeuroD1-expressing microglia under both physiological and disease conditions. Across all approaches, NeuroD1-expressing microglia consistently retained their microglial identity and failed to acquire neuronal markers or morphology. These findings stand in contrast to previous reports claiming NeuroD1-mediated microglia-to-neuron conversion(25, 34), and instead support the notion that earlier observations may have resulted from experimental artifacts such as viral off-target expression(22, 26, 32, 35).

Previous studies have highlighted the challenges in validating direct glia-to-neuron reprogramming(26, 32). While initial reports demonstrated that the overexpression of single transcription factors such as SOX2, ASCL1, or NeuroD1 could convert glial cells into neurons, subsequent investigations using rigorous lineage tracing or live imaging failed to reproduce some of these inter-lineage conversions(21, 33). Our results reinforce these concerns by showing that even in a brain injury environment, NeuroD1 is insufficient to drive microglial fate conversion. In fact, our long-time observation revealed the disappearance of NeuroD1-expressing microglia over time, consistent with our prior finding that NeuroD1 overexpression induces BCL2-dependent apoptosis in microglia(32). Thus, rather than promoting neuronal reprogramming, NeuroD1 expression appears to compromise microglial survival. The combined evidence of LC3/P62 accumulation and increased TUNEL positivity, together with longitudinal in vivo imaging and prior in vitro data, indicates that NeuroD1 overexpression compromises microglial homeostasis and promotes cell death rather than inducing microglia-to-neuron conversion. This may also explain why the reporter efficiency of NeuroD1MG is lower than that of Ai14MG (quantification data not shown). These results suggest that NeuroD1 promotes microglial loss through multiple stress pathways rather than facilitating neuronal reprogramming.

Importantly, our data also demonstrate that NeuroD1 expression in microglia does not provide functional benefits in the TBI model. Behavioral assessments, including grip strength, and rotarod tests, revealed no improvement in motor function, which correlates with the absence of microglia-to-neuron conversion observed histologically. This discrepancy with previous claims that NeuroD1 promotes functional recovery following stroke or injury likely reflects differences in experimental design, particularly the reliance on viral delivery systems prone to off-target effects(34, 36). Our use of a virus-free genetic strategy ensures that NeuroD1 expression is restricted to microglia, thereby providing a more definitive evaluation of its reprogramming capacity. Although microglia may not represent a suitable cellular source for neuronal conversion, our team has developed an alternative therapeutic strategy based on microglia replacement(37, 38). Notably, this approach has been successfully translated into clinical practice and was shown to halt disease progression in patients with adult-onset leukoencephalopathy with axonal spheroids and pigmented glia (ALSP), supporting microglia as a promising target for cell-based therapies in neurological diseases(39).

Together, these findings underscore the necessity of applying stringent lineage-tracing approaches and multimodal validation strategies when evaluating cell fate conversion. While glia-to-neuron reprogramming remains a promising avenue for regenerative medicine, our results caution against premature translation of NeuroD1-based approaches into clinical settings. The lack of microglia-to-neuron conversion observed here does not diminish the therapeutic potential of targeting microglia in brain disorders; rather, it highlights the importance of exploring alternative strategies that preserve or modulate microglial functions without attempting to force lineage conversion. We agree the potential that NeuroD1 is able to induce stem cells or stem-like cells conversion(40), but direct microglia-to-neuron conversion is not applicable. Future studies should focus on identifying context-dependent mechanisms that may enable or restrict glial plasticity, and on developing approaches that combine genetic, epigenetic, and environmental cues to achieve reliable and functional neuronal reprogramming.

Materials and Methods

Animals

Cx3cr1-CreER[B6.129P2(C)-Cx3cr1tm2.1(cre/ERT2)Jung/J, Stock No: 020940], Ai14 [B6.Cg-Gt(ROSA)26Sortm14(CAG-tdTomato)Hze/J, Stock No: 007914], LSL-Neurod1-GFP [Gt(ROSA)26Sortm1(Neurod1-EGFP)Able/BfriJ, Stock No: 034697] mice were purchased from The Jackson Laboratory. CX3CR1-CreER mice were crossed with LSL-Neurod1-EGFP mice to obtain Cx3cr1-CreER+/-::LSL-Neurod1-EGFP+/+ (NeuroD1MG) mice. Similarly, Cx3cr1-CreER mice were crossed with Ai14 mice to obtain Cx3cr1-CreER::Ai14 (Ai14MG) mice. All animals were housed in the Animal Facility of the Department of Laboratory Animal Science at Fudan University on a 12-h light‒dark cycle with food and water provided ad libitum. No significant sex differences were observed in this study.

Drug administration

For the NeuroD1MG and Ai14MG mice, tamoxifen at 150 mg/kg body weight was administered via intragastric gavage (i.g.) for four consecutive days to induce loxP-dependent recombination.

Tissue preparation for immunohistochemistry

The animals were deeply anesthetized with 1% pentobarbital sodium (60 mg/kg) via intraperitoneal injection and transcranially perfused with cold 0.01 M PBS and cold 4% paraformaldehyde (PFA). The brains were immediately separated from the skulls and postfixed in 4% PFA for 24h at 4°C. For vibratome sectioning, brains were directly cut into coronal sections at the desired thickness using a Leica vibratome after post-fixation. For cryo-sectioning, brains were dehydrated in 30% sucrose in 0.01 M PBS at 4 °C for 2–3 days, followed by immersion in fresh 30% sucrose for an additional day. Tissue sections containing the regions of interest were then cut at 35 or 40 μm thickness using a Leica CM1950 cryostat.

Immunohistochemistry

The brain sections were rinsed 3 times in 0.01 M PBS for 10 min. The samples were blocked with 0.01 M PBS containing 0.3% Triton X-100 and 4% normal donkey serum (NDS, Jackson ImmunoResearch, Cat#: 017-000-121) at room temperature (RT) for 2 h. The samples were subsequently incubated with primary antibodies in 0.3% PBST containing 1% NDS at RT overnight. The samples were rinsed 3 times with 0.03% PBST for 10 min and then incubated with fluorescent dye-conjugated secondary antibodies in 0.03% PBST containing 1% NDS at RT for 2 h, together with 1:500 4’,6-diamidino-2-phenylindole (DAPI, Sigma‒Aldrich, Cat#: D9542-10MG). Finally, the samples were rinsed with 0.03% PBST for 10 min 3 times, and the slides were mounted with antifade mounting medium. The edges of the cover glass were then sealed with nail enamel.

The primary antibodies used in this study were as follows: rabbit anti-Iba1 (Wako, Cat#: 019-19741, 1:500), goat anti-IBA1 (Abcam, Cat#: ab5076, 1:500), goat anti-GFP (Abcam, Cat#: AB6673, 1:500), goat anti-mCherry (Biorbyt, cat#: orb11618, 1:500), rabbit anti-GFP (Invitrogen, Cat#: A-11122, 1:1000), mouse anti-NEUN (Abcam, Cat#: ab104224, 1:500), rabbit anti-NeuroD1 (Abcam, Cat#: ab109224, 1:500), rabbit anti-RFP (Abcam, Cat#: ab62341, 1:1000), rabbit anti-DCX (Abcam, Cat#: ab18723, 1:200), mouse anti-MAP2 (Sigma, Cat#M9942: a, 1:2000), and rabbit anti-TUJ1 (Sigma, Cat#: T3952, 1:2000), mouse anti-LC3B (Cell Signaling Technology, Cat#: 83506T, 1:200); rabbit anti-p62 (Cell Signaling Technology, Cat#: 23214T, 1:500). The secondary antibodies used were as follows: AF488 donkey anti-goat IgG (Jackson ImmunoResearch, Cat#: 705-545-003, 1:2000), AF488 donkey anti-rabbit IgG (Jackson ImmunoResearch, Cat#: 711-545-152, 1:2000), AF568 donkey anti-rabbit IgG (Invitrogen by Thermo Fisher Scientific, Cat#: A10042, 1:2000), AF568 donkey anti-goat IgG (Invitrogen by Thermo Fisher Scientific, Cat#: A11057, 1:2000), AF647 donkey anti-rabbit IgG (Jackson ImmunoResearch, Cat#: 711-605-152, 1:2000), AF647 donkey anti-mouse IgG (Jackson ImmunoResearch, Cat#: 715-605-151, 1:2000), and AF647 donkey anti-goat IgG (Jackson ImmunoResearch, Cat#: 705-605-003, 1:2000).

In situ apoptosis assay

A commercial Tunnel Cell Apoptosis Detection Kit (Service bio, Cat#: G1502-50T; Elab science, Cat#: E-CK-A321) was used to detect in situ cell apoptosis. Briefly, brain sections were treated with Proteinase K (20 ug/mL) for 10 mins at RT. Then, brain sections were incubated with equilibration buffer for 10 minutes at RT. Next, apoptotic cells were labeled by TMR-5-dUTP or FITC-dUTP labeling mix at RT and counterstained with DAPI. Positive controls were treated with DNase I and negative controls were labeled by buffer without recombinant TdT enzyme.

Confocal microscopy

Confocal images of the fluorescent samples were acquired via 20x (NA 0.75), 40x (NA 1.30) or 60x (NA 1.40) objectives of a Nikon A2 confocal microscope. The Z-stacked focal planes were taken and maximally projected by ImageJ, whereas brightness and contrast were adjusted in ImageJ if necessary.

Tile scans of whole coronal brain sections were captured with an Olympus VS120 or VS200 system and exported from OlyVIA Software in TIFF format. Further data analysis was carried out in ImageJ.

Two-photon surgery and imaging

Two-photon cranial window surgery was performed on mice anesthetized with isoflurane (3% induction, 1.2%-1.5% maintenance). A 3.5 mm-diameter circular cranial window was created at the skull above the sensory cortex, and a same-diameter circular D263T glass slice was implanted. The cranial window and a head bar were bonded using glue and dental cement to maintain stability. In addition to performing routine postoperative care, the mice were injected intraperitoneally with antibiotics and anti-inflammatory drugs (dexamethasone sodium phosphate injection and ceftiofur sodium for injection) for five consecutive days.

Two-photon imaging was performed on days 18, 22, 26, and 30 post-tamoxifen induction (as shown in the timeline in Fig. 2). During imaging, the mice were anesthetized with isoflurane (3% induction, 1.2%-1.5% maintenance), and body temperature was maintained by a heating pad. Imaging was performed via a self-developed two-photon microscopy system (developed by the Institute for Translational Brain Research, Fudan University) with one resonant and one galvo scan mirror. The excitation was delivered with a Coherent Chameleon Ti-Sapphire pulsed laser. The fluorescence emission was filtered with a Semrork FF01-520/35-25, FF01-593/46 filter and detected with a Hamamatsu photomultiplier tube (H7422-40P). The EGFP signal in Cx3cr1-CreER::LSL-Neurod1-EGFP mice was excited by a 920 nm laser and imaged via a 16x, 0.8 NA water-dipping objective (Nikon). Images were acquired at a depth of approximately 100 μm below the dura using ScanImage(41) at step intervals of 2 μm. Each Z-axis plane image (512×512 pixels) was imaged at a consistent zoom factor (∼300 × 300 μm area) and acquired at a 1.06 Hz frame rate.

Image processing was performed via ImageJ, and the Z-stack image containing the target cells was selected. Brightness and contrast were adjusted if needed, and the maximal projection was performed. Video processing was performed via Adobe Premiere software.

AAV vectors, lentivirus preparation, and stereotaxic injection

The AAV vectors were purchased from Obio Technology, Ltd., and PackGene Biotech, Inc., and were as follows: AAV1/2-CMV-EGFP, AAV5/2-CMV-EGFP, AAV6/2-CMV-EGFP, AAV8/2-CMV-EGFP, AAV9/2-CMV-EGFP, AAV2-DJ-CMV-EGFP, AAV2-retro-CMV-EGFP, AAV-PHP.eB-CMV-EGFP, AAV6-CX3CR1-GFP, AAV8-CX3CR1-GFP, AAV9-CX3CR1-GFP, AAV6-CD68-GFP, AAV8-CD68-GFP, and AAV9-CD68-GFP.

All lentiviruses were packaged by Obio Technology, Ltd., according to the author’s design and requirements. All lentiviruses were packaged in HEK293 cells, and the packaged plasmids consisted of pMDL, VSV-G and pREV. The following lentiviral vectors were used: pLenti-CAG-flag-NeuorD1, pLenti-CAG-flag-MCS, pLenti-CMV-GFP, pLenti-CX3CR1-GFP and pLenti-CD68-GFP.

For intracranial injection, the mouse was anesthetized with isoflurane. The stereotaxic injection coordinates were +1.0 mm anterior/posterior, ± 1.7 mm medial/lateral, and −0.75 mm dorsal/ventral.

Traumatic brain injury (TBI)

TBI was induced using the controlled cortical impact (CCI) model as previously described(42). Briefly, mice were anesthetized with ketamine (100 mg/kg) and xylazine (10 mg/kg) and placed in a stereotaxic frame (Stoelting, Wood Dale, IL, USA). Body temperature was maintained at 37.0 ± 0.5 °C using a heating pad throughout the procedure. After scalp disinfection, a midline incision (1.5–2 cm) was made to expose the skull. A 4 mm craniotomy was performed over the right parietal cortex (between bregma and lambda, 1 mm lateral to the midline) using a portable drill, taking care to preserve the dura mater. Mice with damaged dura were excluded from further analysis. Cortical impact was delivered using a 3 mm rounded steel impactor (PinPoint PCI3000, Hatteras Instruments Inc., USA) at a velocity of 1.5 m/s, impact depth of 1.5 mm, and dwell time of 100 ms. After impact, the bone flap was sealed with sterile bone wax and the scalp was sutured. Sham-operated mice underwent the same surgical procedures without cortical impact. Following surgery, mice were placed in a 37 °C recovery chamber until fully ambulatory.

Rotarod test

The mice were acclimated in the behavioral room for half an hour before testing. Briefly, during the training phase, the mice were trained to walk on a rotating rod at 5 RPM, allowing them to maintain balance by walking forward. Each mouse was placed on the rod for 60 seconds, after which it was returned to its home cage. This procedure was repeated 3 times with a 10-min interval. In the testing phase, the mice were placed in separate lanes on the rod, which was set to accelerate from 4 to 40 RPM over a 300-second period. This procedure was repeated three times at approximately 30-min intervals. The latency to fall was recorded for statistical analysis.

Grip strength test

The mice were housed in the behavioral room half an hour before the behavioral testing. Briefly, the grip strength meter was positioned on the table. Each mouse was gently held by the tail and placed with all the paws on the grip strength grid. The mouse was then gently pulled back until its grasp was released, after which the maximum holding force was recorded. This procedure was repeated 6 times, and the maximum force was used for statistical analysis.

Single-cell preparation and fluorescence-activated cell sorting (FACS)

For scRNA-seq, FACS was used to harvest GFP+ or tdTomato+ cells from tamoxifen-treated Cx3cr1-creER::LSL-Neurod1-GFP and Cx3cr1-CreER::Ai14 mice. Briefly, brains were dissected and cut into 1-mm³ pieces using a mouse stainless steel brain matrix (RWD). The tissue pieces were digested in 3 ml of DMEM containing 8 U/ml papain (Sangon Biotech, cat: A501612), 125 U/ml DNase I (Sangon Biotech, cat: A510099), and 1.5 mg/ml trypsin inhibitor (Aladdin, cat: A274384) via a gentleMACS Octo Dissociator (Miltenyi) with the 37C_ABDK program.

After 30 min of digestion, the mixture was gently pipetted up and down and then filtered through a 70 µm cell strainer. The dissociated cells were centrifuged at 370 × g at RT for 10 min, and the supernatant was discarded. The cell pellets were resuspended in 4 ml of 30% Percoll (Cytiva, cat: 17089101) and centrifuged again at 370 × g and 17°C for 10 min to remove debris. After centrifugation, the supernatant was discarded, and the cells were washed once with DPBS containing 0.5% BSA. Before the cells were loaded onto the SONY MA900, they were stained with 7-AAD (BD Pharmingen, cat: 559925, 1:700) to label dead cells. Finally, the tdTomato+ 7-AAD- or GFP+7-AAD- cells were collected for scRNA-seq immediately.

scRNA-seq data analysis

scRNA-seq was conducted via the MobiCube High-throughput Single Cell 3′ Transcriptome Set V2.1 (PN-S050200301) and the MobiNova-100 microfluidic platform. The single-cell suspension was adjusted to a concentration of 700–1200 cells/μl and immediately loaded onto a chip for microdroplet formation via the MobiNova-100 system. Reverse transcription, cDNA amplification, and DNA library construction were carried out according to the manufacturer’s protocol. High-throughput sequencing was performed in PE-150 mode.

FASTQ files were processed and aligned to the mouse reference genome (mm10) via MobiVision version 3.2, with unique molecular identifier (UMI) counts aggregated for each barcode. The resulting UMI count matrix was analyzed in R version 4.2.2 via Seurat version 4.3.0. To ensure data quality, we excluded low-quality cells and potential multiplets on the basis of the following criteria: cells with fewer than 200 genes or more than 5000 genes and cells with more than 10% UMIs mapped to mitochondrial genes. scRNA-seq data from the four groups were integrated via IntegratedData functions according to the manufacturer’s instructions. The remaining data were normalized via the NormalizeData function and scaled with ScaleData. For scRNA-seq analysis, principal component analysis (PCA) was conducted on variable features, with the top 30 principal components (PCs) selected for clustering and visualization via tSNE and uniform manifold approximation and projection (UMAP). Cell clustering was performed with a resolution of 1 to annotate the cell types. Differentially expressed genes (DEGs) among cell groups were identified via the FindMarkers function with parameters set to min.pct = 0.1 and a log fold change (FC) threshold of 0.25 (Wilcoxon rank-sum test). Genes with |log2FC| > 0.25 and P < 0.05 were considered significant DEGs and subsequently used for Gene Ontology (GO) analysis. Visualization of the results was performed via the VlnPlot and FeaturePlot functions.

Statistical analysis

Statistical analyses were performed with Prism 8.4.0 (GraphPad). Each data point represents the average statistical result of more than three brain sections in different bregma regions of one mouse. The results were evaluated independently in a double-blind manner. All the results are presented as the means ± standard deviations (SDs). One-way analysis of variance (ANOVA) with Holm‒Sidak’s multiple comparisons test (post hoc) was performed for multiple comparisons, whereas a two-tailed independent t test was used to compare the differences between two groups. No outliers were excluded.

eLife Assessment

This is a valuable study addressing a debated question in brain repair research by testing whether NeuroD1 can convert brain immune cells into nerve cells using a virus-free genetic approach and live imaging. The solid evidence presented here supports that, under the conditions tested, NeuroD1-expressing cells do not become nerve cells and instead retain their original identity; however, some aspects of expression level, injury timing, and cell-type composition would benefit from further clarification. This work will be of interest to neuro-immunologists.

Reviewer #1 (Public review):

Summary:

This study revisits an important and controversial question in brain repair: whether NeuroD1 can convert brain immune cells into nerve cells in vivo. Using a virus-free genetic system, in vivo imaging, injury experiments, and single-cell profiling, the authors provide convincing evidence that NeuroD1-expressing cells do not become nerve cells under the tested conditions. Instead, these cells largely retain their original immune-cell identity, and some appear to undergo cellular stress or loss.

Strengths:

The main strength of the work is that it tests this question with a cleaner genetic strategy, avoiding some of the concerns associated with viral delivery and unintended cell labeling. Although the overall conclusion is consistent with the authors' previous work, the current study adds useful independent evidence, particularly through the virus-free fate-mapping system and live imaging in the brain.

Weaknesses:

There are some limitations. In the injury experiment, the labeled cells may include both resident brain immune cells and blood-derived immune cells recruited after injury, so the authors should be cautious when referring to all labeled cells as microglia. The level of NeuroD1 expression achieved by the genetic system is also not fully defined, which matters because the effects of such a cell-fate regulator may depend on expression level. Finally, the tested time window may not fully address very delayed or incomplete neuronal differentiation.

Overall, this is a useful and careful study that supports the conclusion that NeuroD1 does not drive brain immune cells to become nerve cells in the tested settings. It should be valuable for researchers studying brain repair, cell fate conversion, and genetic fate mapping, and it provides a clear caution against overinterpreting reprogramming results based only on viral labeling.

Reviewer #2 (Public review):

Summary:

In vivo glia-to-neuron conversion emerges as a potential regeneration-based therapeutic strategy for neural injuries and diseases. However, controversies exist in this exciting field, largely arising from the non-stringent methods employed for analyzing in vivo neuronal conversions. The study by Li et al. directly addressed this controversy regarding Neurod1-mediated microglia-to-neuron conversion. They took advantage of two transgenic mouse lines to specifically express Neurod1 in the microglia of adult mouse brains. Results from immunohistochemistry, in vivo live-cell imaging, and scRNA-seq convincingly demonstrate that microglia cannot be converted in vivo to neurons by ectopic Neurod1 expression under both normal and injury conditions. Instead, it induces microglia death, consistent with their earlier findings. These solid results, though negative, are critical additions to the field and further support that stringent lineage tracing methods are essential for studying in vivo cell reprogramming. Overall, the studies are rigorously designed and executed. Only minor issues need to be dealt with.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This study revisits an important and controversial question in brain repair: whether NeuroD1 can convert brain immune cells into nerve cells in vivo. Using a virus-free genetic system, in vivo imaging, injury experiments, and single-cell profiling, the authors provide convincing evidence that NeuroD1-expressing cells do not become nerve cells under the tested conditions. Instead, these cells largely retain their original immune-cell identity, and some appear to undergo cellular stress or loss.

Strengths:

The main strength of the work is that it tests this question with a cleaner genetic strategy, avoiding some of the concerns associated with viral delivery and unintended cell labeling. Although the overall conclusion is consistent with the authors' previous work, the current study adds useful independent evidence, particularly through the virus-free fate-mapping system and live imaging in the brain.

Weaknesses:

There are some limitations. In the injury experiment, the labeled cells may include both resident brain immune cells and blood-derived immune cells recruited after injury, so the authors should be cautious when referring to all labeled cells as microglia. The level of NeuroD1 expression achieved by the genetic system is also not fully defined, which matters because the effects of such a cell-fate regulator may depend on expression level. Finally, the tested time window may not fully address very delayed or incomplete neuronal differentiation.

Overall, this is a useful and careful study that supports the conclusion that NeuroD1 does not drive brain immune cells to become nerve cells in the tested settings. It should be valuable for researchers studying brain repair, cell fate conversion, and genetic fate mapping, and it provides a clear caution against overinterpreting reprogramming results based only on viral labeling.

We’d like to thank the reviewer for providing these thoughtful suggestions and positive assessment of our study. We agree that CX3CR1 lineage can label both microglia and blood-derived immune cells after injury, To more specifically assess microglial lineage tracing and minimize the potential contribution of infiltrating peripheral myeloid cells, we performed additional experiments using TMEM119-CreER::LSL-NeuroD1-EGFP mice. These experiments showed no evidence of microglia-to-neuron conversion following TBI. To evaluate the expression level of NeuroD1, we performed qPCR to verify the NeuroD1 expression level in the CD11b+ cell population of the NeuroD1-expressing mice, and found that NeuroD1 expression was increased approximately 6.21-fold compared with control mice, indicating robust NeuroD1 expression. For the question about the time window, we have revised our Discussion and Conclusion parts to avoid overgeneralizing our findings beyond the experimental conditions tested. We agree that our data cannot fully exclude the possibility of delayed neuronal differentiation or neuronal conversion under other pathological conditions. Besides, we have also revised the manuscript accordingly to address the concerns raised by the reviewer.

Reviewer #2 (Public review):

Summary:

In vivo glia-to-neuron conversion emerges as a potential regeneration-based therapeutic strategy for neural injuries and diseases. However, controversies exist in this exciting field, largely arising from the non-stringent methods employed for analyzing in vivo neuronal conversions. The study by Li et al. directly addressed this controversy regarding Neurod1-mediated microglia-to-neuron conversion. They took advantage of two transgenic mouse lines to specifically express Neurod1 in the microglia of adult mouse brains. Results from immunohistochemistry, in vivo live-cell imaging, and scRNAseq convincingly demonstrate that microglia cannot be converted in vivo to neurons by ectopic Neurod1 expression under both normal and injury conditions. Instead, it induces microglia death, consistent with their earlier findings. These solid results, though negative, are critical additions to the field and further support that stringent lineage tracing methods are essential for studying in vivo cell reprogramming. Overall, the studies are rigorously designed and executed. Only minor issues need to be dealt with.

We’d like to thank the reviewer for providing these thoughtful comments and for the positive assessment of our study. We are glad that the reviewer agrees that our results from immunohistochemistry, in vivo live-cell imaging, and scRNA-seq support the conclusion that microglia cannot be converted into neurons by NeuroD1 expression under the tested conditions. We have carefully addressed the minor issues raised by the reviewer, including rechecking the grammar throughout the manuscript, revising the description of the TBI behavioral results, discussing the limitations of scRNA-seq for neuronal detection, adding the information of the promoters used in the study, and carefully revising the references. Besides, we have carefully revised the manuscript according to the reviewer’s suggestions.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Major comments

(1) TBI lineage tracing may label both microglia and infiltrating macrophages. Because tamoxifen was administered after TBI, CX3CR1-CreER may label not only resident microglia but also injuryrecruited CX3CR1+ monocytes/macrophages. The authors should avoid referring to all reporterpositive cells as microglia unless this is clearly supported. They should clarify the composition of reporter-positive cells in the TBI setting, ideally using existing scRNA-seq data or markers distinguishing microglia from infiltrating macrophages.

The current data still support the conclusion that CX3CR1-lineage myeloid cells do not show obvious neuronal conversion in this TBI model, but the wording should be more precise.

Thank you for pointing this out. We agree that, because TBI can recruit CX3CR1-expressing monocytes/macrophages, CX3CR1-CreER lineage tracing after injury may label both resident microglia and infiltrating myeloid cells. To more specifically assess the contribution of resident microglia, we therefore performed additional lineage-tracing experiments using TMEM119CreER::Ai14 (TMEM119-Ai14) and TMEM119-CreER::LSL-NeuroD1-EGFP(TMEM119-ND1) mice, with tamoxifen administration before or after TBI. In both experimental settings, we did not detect reporter-positive cells co-expressing the neuronal marker NeuN in either the lesion core or distal regions. These results provide no evidence of microglia-to-neuron conversion in the TBI model within the examined time window. We have also revised the text throughout the manuscript to distinguish resident microglia from broader CX3CR1-lineage myeloid cells where appropriate.

TMEM119-CreER::LSL-NeuroD1-EGFP animal also show no microglia-to-neuron conversion in TBI.

(A)Tamoxifen induction after TBI shows there is no GFP+NeuN+ cells in both lesion core and distal region. (B)Tamoxifen induction before TBI shows there is no GFP+NeuN+ cells in both lesion core and distal region.

(2) NeuroD1 expression level should be better characterized. The authors show that GFP-positive cells co-express NeuroD1, but the approximate NeuroD1 expression level is not clear. This is important because the effect of a fate-determining transcription factor may be dose-dependent. The authors should provide, if possible, a quantitative estimate of NeuroD1 expression using available IF, scRNA-seq, qPCR, or other data. This would help interpret the negative reprogramming result and would also be useful for future studies using this Rosa26-LSL-NeuroD1-IRES-GFP mouse line.

Thank you for this valuable suggestion. To quantitatively assess NeuroD1 expression in the myeloid compartment, we isolated CD11b+ cells from CX3CR1-CreER::Ai14 (CX3CR1-Ai14) and CX3CR1-CreER::LSL-NeuroD1-EGFP (CX3CR1-ND1) mice using MACS and performed qPCR analysis of Neurod1 expression. Neurod1 expression was increased approximately 6.21-fold in CD11b+ cells from CX3CR1-ND1 mice compared with the corresponding control mice (Author response image 2). These results confirm robust NeuroD1 expression in the CD11b+ cell population of the NeuroD1-expressing mice. This level of induction is broadly comparable to the approximately 4-fold increase in Neurod1 expression reported following NeuroD1 induction in the intestine, in which the same LSL-NeuroD1-EGFP transgenic mouse line was used [1].

Using qPCR verify the relative expression of Neurod1.

(3) Figure 4 scRNA-seq should show Neurod1 expression. The scRNA-seq data show that reporterpositive cells retain microglial markers and lack neuronal markers, but Neurod1 expression itself is not shown. The authors should consider adding Neurod1 feature plots, violin plots, or average expression comparisons between control and NeuroD1-expressing groups. This would help validate the expression system and estimate the level achieved in sorted reporter-positive cells.

Thank you for your suggestion. We examined the expression of Neurod1 in our scRNA-seq database, but the Neurod1 transcripts were rarely detected (not just in this database, also our previous dataset). This might because of Neurod1 is a transcription factor and single-cell RNA-seq method is not suitable for detect these low-expression transcript factor genes, so we utilized qPCR to evaluate the expression of Neurod1, as shown in Author response image 2.

(4) The microglial loss phenotype may be related to NeuroD1/EGFP dosage. In Figure 3, the loss of GFP-positive microglia may reflect a specific effect of NeuroD1 in microglia, but it could also be related to excessive transgene expression or expression-system toxicity. The authors should discuss this possibility and, if possible, examine whether stress/death markers correlate with NeuroD1 or GFP expression intensity.

Thanks for your suggestion. We agree with that, so we performed TUNEL staining on CX3CR1-CreER::Ai14 (CX3CR1-Ai14 for short) and CX3CR1-CreER::LSL-NeuroD1EGFP(CX3CR1-ND1 for short) mice at D4 and D18. The data shows that TUNEL+ Reporter+ cells were higher in NeuroD1-expressing microglia, so we’d like to say death cells (or apoptotic cells) are positively correlated with NeuroD1 expression. Besides, we also discussed more about this. See Figure 3 E, F, G and Fig. S3E

(5) The time window and injury context should be more cautiously discussed. The study follows the genetic model up to 45 days. This is informative, but delayed or incomplete neuronal differentiation cannot be fully excluded, especially if conversion would require a longer repair phase after injury. In addition, except for TBI, the study does not include another neuronal-loss context that might provide a permissive regenerative niche. The authors do not necessarily need additional long-term experiments, but they should avoid overgeneralizing beyond the tested time window and injury model.

Thank you for this thoughtful comment. We agree that our study evaluated the effects of NeuroD1 expression within a defined experimental time window (up to 45 days) and under the specific physiological and injury conditions examined. Therefore, our data cannot exclude the possibility that neuronal conversion might occur at later time points or under other pathological conditions that provide a more permissive regenerative environment. In response to the reviewer's suggestion, we have revised the Discussion and Conclusion to avoid overgeneralizing our findings beyond the experimental conditions tested. We now emphasize that NeuroD1 expression alone did not induce microglia-to-neuron conversion within the time frame and injury paradigms examined in this study, rather than concluding that such conversion can never occur under other conditions.

Minor comments

(1) In Figure 2, two panels are labeled "D"; the second should likely be "E."

Thank you for pointing this out, we have corrected it in figure legend.

(2) Several spelling and grammar errors should be corrected, such as "Represent images," "illutrating," and "GFP-postive."

Thank you for pointing this out, we have corrected it in our manuscript.

(3) More scRNA-seq details should be provided, including cell numbers per group, QC metrics, and cluster annotation information.

Thanks for your suggestion, we have added more details about scRNA-seq data as shown in Figure S4.

(4) The authors should maintain cautious wording throughout, especially when referring to "microglia" versus broader CX3CR1-lineage myeloid cells.

Thanks for your suggestion, we added the data about we utilize TMEM119-ND1 mouse line to trace microglia-to-neuron conversion, which is specific to microglia not myeloid cell. Besides, we also carefully revise these minor comments in our manuscript.

Reviewer #2 (Recommendations for the authors):

The following are some required minor edits.

(1) Please recheck the grammar throughout the manuscript.

Thank you for pointing this out, we have carefully revised our manuscript.

(2) Figure 4B-C: "indicating that NeuroD1 expression in microglia does not rescue TBI induced motor dysfunction". This should be rephrased, since dysfunction was not detected when comparing TBI and sham.

Thank you for pointing this out, we changed our expression into “indicating that NeuroD1 expression in microglia does not improve the performance in TBI model”.

(3) Interpretation of the scRNA-seq data needs to be cautious, since neurons normally do not survive well during this procedure.

Thank you for your suggestion, we discussed about this in our discussion part.

(4) It will be informative to list the sources or sequences of the promoters used.

Thanks for your advice, all the virus information are provided in our method part.

(5) Some of the references are repetitive.

Thank you for your suggestion, we have carefully revised the reference.

Reference

(1) Li, H. J. et al. Intestinal Neurod1 expression impairs paneth cell differentiation and promotes enteroendocrine lineage specification. Sci Rep 9, 19489 (2019). https://doi.org/10.1038/s41598-019-55292-7

Version 2

Abstract

In situ glia-to-neuron conversion induced by a single transcription factor is a promising regenerative strategy for treating central nervous system injuries and neurodegenerative diseases. Previous studies reported that NeuroD1 can induce microglia-to-neuron cross-lineage conversion and improve recovery after brain injury. However, these findings remain controversial. To examine the ability of NeuroD1 in inducing microglia-to- neurons conversion, we employed virus-free genetic lineage tracing systems to specifically induce NeuroD1 expression in microglia, and track the cell fate of NeuroD1-expressing cells at multiple timepoints. Meanwhile, we longitudinally monitored NeuroD1-expressing microglia via two-photon imaging, and characterized their transcriptional profile by single-cell RNA sequencing. All observations across these methods revealed that NeuroD1-expressing cells retained their microglia identity, and cannot convert into neurons, no matter under physical and injury conditions. Instead, sustained expression of NeuroD1 facilitated microglia apoptosis in vivo. Collectively, our findings provide strong evidence that NeuroD1 alone is sufficient to induce microglia-to neuron conversion. This study further highlighted the importance of rigorous lineage-tracing and cell fate mapping strategies for validating in situ glia-to-neuron conversion.

Introduction

In the central nervous system (CNS), neurons have limited regenerative capacity, therefore, brain injury and neurodegeneration result in permanent neuronal loss. In contrast, glia cells, especially microglia exhibit a high regenerative capacity(1–3). In situ reprogramming glial cells into neurons offers a promising strategy for neuronal regeneration and functional recovery in brain injuries and neurodegenerative disease. Recent years, single gene such as NeuroD1 (4), PAX6 (5), SOX2 (6–8), ASCL1 (9) and PTBP1 (10, 11) have been reported to induce inter-neuroectodermal lineage conversion, transforming astrocytes, oligodendrocyte precursor cells (OPCs) or Muller glia into neurons, shedding a light for neurodegenerative therapies.

However, before clinical translation, these approaches require rigorous validation using stringent and robust approaches. Unbaised and well-controlled lineage tracing, and unambiguous live imaging are regard as gold standards for validating glia-to-neuron conversion (12, 13). Indeed, several studies employing rigorous lineage tracing and live imaging techniques have failed to observe some inter-lineage glia-to-neuron conversions with single-gene manipulation (14–22), including NeuroD1 (20–22). In addition to inter-lineage conversion, Nakashima et al. have reported that NeuroD1 can induced microglia-to-neuron cross-lineage conversion (23). We previously conducted lineage tracing using virus-based tracing system, in vitro live-cell imaging, and microglia ablation, and demonstrated that the reported cross-lineage conversion of microglia into neurons may be attribute to experimental artifacts, e.g., off-target transduction from virus (13). Although Nakashima et al. argued that the phenomenon we observed was due to insufficient NeuroD1 expression in microglia (24), our comparison of the data from both studies (13, 23) indicates that our NeuroD1 expression levels are comparable to, or even higher than, those reported in their study (25). Besides, Nakashima et al. reported that NeuroD1 could induce the microglia-to neurons conversion at brain injury sites in vivo, and promote recovery after brain injury (26). Such conflicting reports are of particular concern, because they directly influence whether glia-to-neuron reprogramming could be translated in to clinical. To settle this controversy, we employed virus-free genetic lineage tracing system to avoid potential experimental artifacts, longitudinal two-photon live imaging to direct monitor the changes in NeuroD1-expressing microglia in vivo, and single cell RNA sequencing (scRNA-seq) to characterize their transcriptional profiles. Our findings provide strong evidence that microglia do not undergo conversion into neurons in vivo, under either physiological or pathological conditions. Our study highlighted the importance of employing rigorous lineage tracing, longitudinal live cell imaging, and multimode validation in evaluating glia-to neuron reprogramming approaches.

Results

Genetic lineage tracing reveals NeuroD1 expression in microglia fails to induce neuron conversion in vivo

We previous reported that ectopic expression of NeuroD1 using lentiviral vector does not induce the microglia-to-neurons conversion, but triggers microglial cell death (13). However, most viral tools designed for microglia exhibit limited efficiency and specificity in the brain, but displayed off-target transduction of neurons and astrocytes (13, 20, 27). This limitation was observed in either lentivirus or adeno-associated virus (AAV) (with different serotypes 6, 8 and 9), and irrespective of the use of myeloid CX3CR1 or CD68 promoters (Figure S1). Similar observations have been reported in our previous study and other groups (13, 20, 27).

Direct evidence for glia-to-neuron conversion by NeuroD1 via a virus-free expression system, particularly genetic lineage tracing is lacking. Genetic lineage tracing are essential for avoiding experimental artifacts relate to virus leakage (12, 13). Thus, we utilized transgenic animals to induce NeuroD1 specifically expression in microglia, and track the cell fate of NeuroD1-expressing microglia. We therefore crossed macrophage/microglia-targeting CX3CR1-CreER mouse line, or a microglia-specific TMEM119-CreER mouse line (28, 29), with an inducible NeuroD1-expressing mouse line Rosa26-LoxP-Stop-LoxP-NeuroD1-IRES-GFP (LSL-NeuroD1-GFP) (30), to obtain CX3CR1-CreER::LSL-NeuroD1-GFP mice (hereafter, CX3CR1-ND1) and TMEM119-CreER::LSL-NeuroD1- GFP mice (hereafter, TMEM119-ND1 for short). Meanwhile, CX3CR1-CreER::Ai14 (CX3CR1-Ai14) and TMEM119-CreER::Ai14 (TMEM119-Ai14) mice were used as controls (Figure 1A and Figure S2A). Following tamoxifen treatment, all GFP positive cells were continuously co-expressed with NeuroD1 (Figure 1B-1D and Figure S2B-2C), enabling us to tracing NeuroD1-expressing cells for lineage tracing. Meanwhile, we observed that theses GFP positive cells retained the characteristics morphology of microglia, exhibiting typical highly ramified process and small cell body (Figure 1 and Figure S2).

Virus-free genetic induction can successfully induce NeuroD1 expression in microglia.

(A) Schematic of the specific expression of NeuroD1, lineage tracing of microglia and experimental timeline. D, day. i.g. oral gavage. (B) Representative confocal images at D4, D10, D30 and D45 showing that NeuroD1 is robustly co-expressed in GFP-positive cells after Cre recombination. (C) Representative confocal images at D18 showing that NeuroD1 is robustly co-expressed in GFP-positive cells after Cre recombination. (D) Quantification of the percentage of NeuroD1+GFP+ cells among GFP+ cells and NeuroD1+GFP- cells among NeuroD1+, revealing NeuroD1 is highly co-expressed with GFP. N = 4-6 mice for each group. Data are presented as mean ± SD.

We subsequently examined the fate of GFP positive cells at multiple timepoints, at 4, 10, 18, 30, and 45 days post–tamoxifen induction. Immunostaining for microglia and neuron markers revealed that all GFP positive cells were consistently co-localized with microglia marker IBA1, but they did not express neuronal markers, NeuN, TUJ1 and MAP2 within CX3CR1-ND1 mouse brain (Figure 2 and Figure S3A-3D). The observations were the same in TMEM119-ND1 (Figure S2). These findings indicate that sustained NeuroD1 ectopic expression by genetic modification, microglia retain their microglia identify but not acquire a neuronal cell fate.

Sustained NeuroD1 expression in microglia does not induce microglia-to-neuron conversion across time.

(A) Experimental timeline. D, day. i.g. oral gavage. (B) Representative confocal images at D4, D10, D18, D30 and D45 showing that GFP is co-labeled with IBA1 but not co-labeled with NeuN. (C) Quantification of IBA1+GFP+ double-positive cells among GFP+ cells or NeuN+GFP+ double-positive cells among NeuN+ in cortex at D4, D10, D18, D30 and D45. N = 4 or 5 mice for each group. (D) Quantification of IBA1+GFP+ double-positive cells among GFP+ cells or NeuN+GFP+ double-positive cells among NeuN+ in striatum at D4, D10, D18, D30 and D45. N = 4 or 5 mice for each group.

Ectopic NeuroD1 expression induce microglia apoptosis rather than neuronal conversion

Given the possibility that genetically labeled cells may loss or downregulation of reporter, such as GFP expression during cell fate conversion (31), we further monitored the cell fate of GFP- labeled microglia in vivo using 2-photon microscopy (Figure 3A). We repeated image GFP- positive cells in the frontal cortex with 4 days intervals, beginning on day 18, and continuing throughout day 45 post-tamoxifen administration (Figure 3A). This strategy allowed us to directly track individual NeuroD1-expressing microglia over time. The longitudinal live cell imaging revealed that there was no obvious morphological transition during the imaging period (Figure 3B, white rectangle, and Movie 1). Instead, the GFP-positive cells consistently exhibited a highly ramified morphology with short, and motile processes, which were typical morphological features of microglia rather than neuron (Figure 3B, white rectangle and Movie 1). Meanwhile, we observed that some GFP-labeled microglia progressively disappeared during the imaging period (yellow arrow in Figure 3B, and Movie 1). These observations indicate sustained ectopic NeuroD1 expression could not induce microglia-to-neuron conversion, but induce microglia cell loss in vivo. These in vivo time lapsing imaging observations were consistent with our previous finding from in vitro live cell imaging (13). We further quantified numbers of GFP-labeled microglia at multiple timepoints after tamoxifen induction. The number of GFP-labeled cells initially increased and reached a peak at day 30 post-tamoxifen induction, but subsequently declined by day 45 (Figure 3C, D). The progressive loss of GFP-labeled microglia was also observed in TMEM119-ND1 brain (Figure S2D). These findings further supported that NeuroD1 expression in microglia induce progressively cell loss overtime in vivo.

NeuroD1-expressing microglia retain microglial their characteristic morphology, but undergo progressively cell loss over time.

(A) Experimental timeline. D, day. i.g. oral gavage. (B) Representative images showing GFP-positive cells with highly ramified morphologies characteristic of microglia rather than neurons. The white rectangular indicate no morphological change, the yellow arrow indicating the GFP+ cell loss over time. (C) Representative images illustrating changes in the number of GFP-positive cells over time (green dots indicate GFP-positive cell). (D) Quantification of GFP-positive cells density. N = 4 - 6 mice per time point. Data are presented as mean ± SD. One Way ANOVA with Turkey’s multiple comparison test. (E) Experimental timeline. (F) Representative confocal image shows TUNEL+ Reporter+ cells are increased in CX3CR1-ND1 animal. (G) Quantification of TUNEL+ Reporter+ cells at D4 and D18. Unpaired t test. Data are presented as mean ± SD. (H) Relative gene expression level at D18, Unpaired t test. Data are presented as mean ± SD. (I) Representative confocal image shows the LC3 and P62 level in CX3CR1-Ai14 and CX3CR1-ND1mice. Scale bar, 50μm. (J) Quantification of P62+ and LC3+ area per cell. Unpaired t test. Data are presented as mean ± SD.

To investigate the mechanisms underlying NeuroD1-expressing microglia loss, we first asked that whether the progressive loss of NeuroD1-expressing microglia due to cell apoptosis. To this end, we assessed apoptosis by TUNEL staining in CX3CR1-ND1 and CX3CR1-Ai14 mice (Figure 3E). In line with the gradually disappear of GFP positive cells, we observed that there was a significantly higher proportion of TUNEL positive cells among reporter positive cells in CX3CR1-ND1 mice than that of controls, increasing from approximately 0.01% in controls to 1.7% in NeuroD1-expressing cells (Figure 3F and 3G). This indicate that NeuroD1 expression accelerated microglia apoptosis in vivo. We next asked the molecular mechanisms through which NeuroD1 induce microglia cell death. Microglia were isolated from CX3CR1-ND1 and CX3CR1-Ai14 brains using CD11b-based magnetic separation at day 18 post tamoxifen induction (Figure 3E and Figure S3A). We then performed qPCR to examine the expression of genes involved in apoptosis (Parp1, Casp3, Bax, Bcl2), necrosis (Ripk1, Fas, Mlkl), autophagy (Becn1 and Atg5) (25, 32, 33). Among the genes examined, Atg5, a key regulator of autophagy (34), was significantly upregulated in cells isolated from CX3CR1-ND1 compared with controls (Figure 3H). We therefore examined autophagic markers following NeuroD1 induction. In CX3CR1-ND1 mice, reporter+ cells exhibited about 4-fold higher LC3 and P62 levels than controls, as shown by immunostaining (Figure 3I-J), suggesting accumulation of autophagosomes markers and impaired autophagic flux. Taken together, these data indicate that sustained ectopic NeuroD1 expression induces microglia apoptosis, accompanied with autophagic dysregulation, leading to NeuroD1-expressing microglia loss in vivo, rather than microglia-to-neuron conversion.

NeuroD1 fails to reprogram microglia into neurons even in the injured brain

Previous studies suggest that injury environments might precondition glial cells for fate reprogramming. For example, Heinrich et al. demonstrated that a stab wound prior to virus delivery is necessary to stimulate NG2 glia to neuron conversion by SOX2(35). Similarly, Irie et al. showed that NeuroD1 can directly reprogram microglia/macrophages into neurons at the lesion site following transient middle cerebral artery occlusion(26). These findings raise the possibility that the failure of NeuroD1-mediated microglia-to-neuron conversion under physiological conditions may attribute to the absence of injury-induced reprograming permissive microenvironment.

To test this possibility, we established a moderate traumatic brain injury (TBI) in the motor cortex in CX3CR1-ND1 mice as previously describe (36), followed by tamoxifen induction to activate NeuroD1 expression in microglia (Figure 4A). If the injury environment stimulates NeuroD1-meidiated microglia to neuron conversion, the new born neurons would contribute to motor function recovery after TB1. We therefore performed a serries behavior test to assess motor recovery at 19 days post tamoxifen induction (Figure 4A). In grip strength task, the CX3CR1-ND1 mice showed comparable performance to that of controls, with no significant differences in the strength between groups (Figure 4B). Similarly, no improvements were observed in rotarod task, as reflected by a comparable latency to fall between groups (Figure 4C). These results indicate that NeuroD1 expression in microglia does not improve the motor function recovery in TBI model. We next directly examined the cell fate of NeuroD1-expressing microglia in the injury brain. Immunostaining analysis showed that GFP-labeled NeuroD1-expressing microglia did not co-localize with the neuronal marker NeuN (Figure 4D, E) or MAP2 (Figure 4F, G), both in the lesion core and distal regions. To further figure out the cell fate of microglia after NeuroD1 induction, we conducted scRNA-seq to characterize the transcriptional profile of NeuroD1-expressing cells (Figure 4A). Both tdTomato and GFP positive cells were isolated from CX3CR1-Ai14 and CX3CR1-ND1 brain respectively, using flow cytometry (Figure S4A), under both physiological and TBI condition. These cells were subsequently subjected to scRNA-seq (Figure S4B-4E). In line with the immunofluorescent findings, when we annotated sorted reposter-positve cells, we did not identify any neuronal cluster, even under injury condition (Figure 4H, Figure S4E). Additionally, the reporter-positive cells still strong expressed microglia markers, including P2ry12, Hexb, and Tmem119 (Figure 4I and 4J), but did not express neuronal markers, such as Rbfox3, Tubb3, and Map2 (Figure 4I and 4J). Collectively, these findings demonstrate that cannot enable NeuroD1 fails to induce microglia-to-neuron conversion, even under an injury condition.

NeuroD1 does not induce microglia-to-neuron conversion even in an injury conditioned microenvironment.

(A) Experimental design. D, day. i.g. oral gavage. (B) Quantification of grip strength showing no improvement in NeuroD1-expressing TBI mice. Two-way ANOVA with multiple comparisons. (C) Quantification of rotarod performance showing no improvement in NeuroD1-expressing TBI mice. Two-way ANOVA with multiple comparisons. (D) Representative confocal images of DAPI, GFP and NeuN co-staining results in CX3CR1-ND1 TBI mice, showing no co-localization of GFP with NeuN in either the lesion core or distal regions. Scale bar, 50 μm. (E) Quantification of the percentage of GFP+NeuN+ cells among total NeuN+ cells in both the lesion area and distal area. n.d., not detected. (F) Representative confocal images of DAPI, GFP, and MAP2 in CX3CR1-ND1 mice, showing no co-localization of GFP with MAP2 in either the lesion core or distal regions. Scale bar, 50 μm. (G) Quantification of the percentage of GFP+MAP2+ cells among total MAP2+ cells in the lesion area and distal region. n.d., not detected. (H) UMAP reduction shows the cell types of the FACS-sorted reporter+ cells in each group; note that there are no neurons. (I) Feature plot showing that the microglia-specific marker P2ry12 was highly expressed even under TBI conditions, but the neuron-specific marker Rbfox3 was barely expressed. (J) Violin plot showing the expression of microglia-specific genes (Hexb and Tmem119) and neuron-specific genes (Tubb3 and Map2) in 4 groups. (K) Experiment timeline of TMEM119-ND1 mice, TBI was performed before tamoxifen induction. (L) Representative confocal images of GFP, IBA1 and NeuN in TMEM119-ND1 mice, showing no co-localization of GFP with NeuN in either the lesion core or distal regions. Scale bar, 50 μm. (M) Quantification of the percentage of GFP+NeuN+ cells among total NeuN+ cells in the lesion area and distal region. n.d., not detected. (N) Experiment timeline of TMEM119-ND1 mice, TBI was performed after tamoxifen induction. (O) Representative confocal images of GFP, IBA1 and NeuN in TMEM119-ND1 mice, showing no co-localization of GFP with NeuN in either the lesion core or distal regions. Scale bar, 50 μm. (P) Quantification of the percentage of GFP+NeuN+ cells among total NeuN+ cells in the lesion area and distal region. n.d., not detected.

Given tamoxifen was administrated after TBI induction in CX3CR1-ND1 mice, when circulating monocytes and macrophages were recruited in to injury region, the CX3CR1-CreER system can also label these peripheral myeloid cells during this period (29, 36) To exclude the recruited myeloid cells mis-led the lineage tracing result, we repeated the same experiments in TMEM119-ND1 mice (Figure 4K). The TMEM119-CreER system allows us to restrict NeuroD1 and GFP expression specifically to resident microglia in the brain (28). Consistent with the observations in CX3CR1-ND1 mice, we did not detect any GFP and NeuN co-localized cells in either the lesion core or regions distal to the injury site (Figure 4L). In addition, we reserved the sequency by conducting TBI after tamoxifen treatment, in turn, labeling resident microglia before the recruiting of peripheral myeloid cells (Figure 4N). Similarly, we observed no GFP-positive cells co-expressing pan neuron marker NeuN in this condition as well (Figure 4O). These findings further proved that NeuroD1 does not reprogram resident microglia into neurons even under injury environment.

Discussion

In current study, we employed virus-free genetic lineage tracing system based on CX3CR1-creER and TMEM119-creER to trace the cell fate of NeuronD1 expressing cells, instead of viral based approaches. Through immunostaining assay at multiple timepoints, two-photon live cell imaging and scRNA sequencing, we systematically assessed the cell fate of NeuroD1-expressing microglia under both physiological and injury conditions. Collectively, all concordance observations provide strong evidence that NeuroD1-expressing microglia consistently retained their microglial identity and failed to acquire neuronal cell fate. These findings in current study contrast with previous reports claiming NeuroD1-mediated microglia-to-neuron conversion(24, 26), and instead support the notion that NeuroD1 is insufficient to induce microglia-to-neuron conversion. We speculated that the observed microglia to neuron conversion reported in previous study may have resulted from off-target viral expression, which leading to the mis interpretation of the cell source of the observed neurons (13, 21, 22, 25).

Previous studies have highlighted the controversial observations in validating direct glia-to-neuron reprogramming(13, 25). While several studies demonstrated that the single transcription factors, such as SOX2, ASCL1, or NeuroD1, could convert glial cells into neurons, subsequent investigations using rigorous lineage tracing or live imaging have failed to reproduce some of these inter-lineage conversions(20, 35). In fact, our longitudinal live cell imaging revealed a gradual loss of NeuroD1-expressing microglia over time. These findings were consitent with our previous results that ectopic NeuroD1 facilitated microglia cell apoptosis (13). The gradual disappearance of NeuroD1-expressing microglia overtime, LC3/P62 accumulation and increased TUNEL positivity, collectively indicate that NeuroD1 compromises microglial homeostasis and facilitate cell death rather than induces microglia-to-neuron conversion. Although the present study suggests that in situ microglia-to-neuron conversion remains technically challenging and require novel approaches, our team has developed an alternative therapeutic strategy, that is microglia replacement by bone marrow transplantation (Mr BMT) (37, 38). Mr BMT allows healthy microglia carrying normal or functional genes to replace diseased microglia carrying mutated genes, therefore providing an avenue to cure disease associated with primary microgliopathies (37). Indeed, Mr BMT has been successfully translated into clinical practice and was shown to halt disease progression in patients with adult-onset leukoencephalopathy with axonal spheroids and pigmented glia (ALSP), supporting Mr BMT as a promising cell-based therapies in neurological diseases(39).

Together, these findings underscore the importance of stringent lineage-tracing approaches and multimodal validation strategies when evaluating cell fate conversion. Although glia-to-neuron reprogramming remains a promising avenue for regenerative medicine, accumulating studies, including the present study raise the caution in interpreting in vivo glia-to-neuron conversion that without rigors validation of cellular origin and cell fate transition, particularly when considering the clinical translation of such approaches.

Materials and Methods

Animals

CX3CR1-CreER[B6.129P2(C)-Cx3cr1tm2.1(cre/ERT2)Jung/J, Stock No: 020940], Ai14 [B6.Cg-Gt(ROSA)26Sortm14(CAG-tdTomato)Hze/J, Stock No: 007914], TMEM119-CreER (C57BL/6-Tmem119em1(cre/ERT2)Gfng/J, Stock No: 031820) and LSL-Neurod1-GFP [Gt(ROSA)26Sortm1(Neurod1-EGFP)Able/BfriJ, Stock No: 034697] mice were purchased from The Jackson Laboratory. CX3CR1-CreER mice were crossed with LSL-Neurod1-EGFP mice to obtain CX3CR1-CreER+/CreER::LSL-Neurod1-EGFP+/+ (CX3CR1-ND1) mice. Similarly, CX3CR1-CreER mice were crossed with Ai14 mice to obtain CX3CR1-CreER::Ai14 (CX3CR1-Ai14) mice. All animals were housed in the Animal Facility of the Department of Laboratory Animal Science at Fudan University on a 12-h light‒dark cycle with food and water provided ad libitum. No significant sex differences were observed in this study.

Drug administration

For mice aged 8 weeks or oder, CX3CR1-ND1, CX3CR1-Ai14, TMEM119-ND1 and TMEM119-Ai14 mice, tamoxifen at 150 mg/kg body weight was administered via intragastric gavage (i.g.) or intraperitoneal injection (i.p.) for four consecutive days to induce loxP-dependent recombination. For mice aged 4 to 6 weeks, the dose was reduced to 75 mg/kg body weight with the same adminstration sechdule.

Tissue preparation for immunohistochemistry

The animals were deeply anesthetized with 1% pentobarbital sodium (60 mg/kg) via intraperitoneal injection and transcardially perfused with cold 0.01 M PBS and cold 4% paraformaldehyde (PFA). The brains were immediately separated from the skulls and postfixed in 4% PFA for 24h at 4°C. For vibratome sectioning, brains were directly cut into coronal sections at the desired thickness using a Leica vibratome after post-fixation. For cryo-sectioning, brains were dehydrated in 30% sucrose in 0.01 M PBS at 4 °C for 2–3 days, followed by immersion in fresh 30% sucrose for an additional day. Tissue sections containing the regions of interest were then cut at 35 or 40 μm thickness using a Leica CM1950 cryostat.

Immunohistochemistry

The brain sections were rinsed 3 times in 0.01 M PBS for 10 min. The samples were blocked with 0.01 M PBS containing 0.3% Triton X-100 and 4% normal donkey serum (NDS, Jackson ImmunoResearch, Cat#: 017-000-121) at room temperature (RT) for 2 h. The samples were subsequently incubated with primary antibodies in 0.3% PBST containing 1% NDS at RT overnight. The samples were rinsed 3 times with 0.03% PBST for 10 min and then incubated with fluorescent dye-conjugated secondary antibodies in 0.03% PBST containing 1% NDS at RT for 2 h, together with 1:500 4’,6-diamidino-2-phenylindole (DAPI, Sigma‒Aldrich, Cat#: D9542-10MG). Finally, the samples were rinsed with 0.03% PBST for 10 min 3 times, and the slides were mounted with antifade mounting medium. The edges of the cover glass were then sealed with nail enamel.

In the present study, the following primary antibodies were used, including rabbit anti-Iba1 (Wako, Cat: 019-19741, 1:500), goat anti-IBA1 (Abcam, Cat: ab5076, 1:500), goat anti-GFP (Abcam, Cat: AB6673, 1:500), goat anti-mCherry (Biorbyt, Cat: orb11618, 1:500), rabbit anti-GFP (Invitrogen, Cat: A-11122, 1:1000), mouse anti-NEUN (Abcam, Cat: ab104224, 1:500), rabbit anti-NeuroD1 (Abcam, Cat: ab109224, 1:500), rabbit anti-RFP (Abcam, Cat: ab62341, 1:1000), rabbit anti-DCX (Abcam, Cat: ab18723, 1:200), mouse anti-MAP2 (Sigma, Cat: M9942, 1:2000), and rabbit anti-TUJ1 (Sigma, Cat: T3952, 1:2000), mouse anti-LC3B (Cell Signaling Technology, Cat: 83506T, 1:200), and rabbit anti-p62 (Cell Signaling Technology, Cat: 23214T, 1:500). The secondary antibodies included AF488 donkey anti-goat IgG (Jackson ImmunoResearch, Cat: 705-545-003, 1:2000), AF488 donkey anti-rabbit IgG (Jackson ImmunoResearch, Cat: 711-545-152, 1:2000), AF568 donkey anti-rabbit IgG (Invitrogen by Thermo Fisher Scientific, Cat: A10042, 1:2000), AF568 donkey anti-goat IgG (Invitrogen by Thermo Fisher Scientific, Cat: A11057, 1:2000), AF647 donkey anti-rabbit IgG (Jackson ImmunoResearch, Cat: 711-605-152, 1:2000), AF647 donkey anti-mouse IgG (Jackson ImmunoResearch, Cat: 715-605-151, 1:2000), and AF647 donkey anti-goat IgG (Jackson ImmunoResearch, Cat: 705-605-003, 1:2000).

In situ apoptosis assay

A commercial Tunnel Cell Apoptosis Detection Kit (Service bio, Cat#: G1502-50T; Elab science, Cat#: E-CK-A321) was used to detect in situ cell apoptosis. Briefly, brain sections were treated with Proteinase K (20 ug/mL) for 10 mins at RT. Then, brain sections were incubated with equilibration buffer for 10 minutes at RT. Next, apoptotic cells were labeled by TMR-5-dUTP or FITC-dUTP labeling mix at RT and counterstained with DAPI. Positive controls were treated with DNase I and negative controls were labeled by buffer without recombinant TdT enzyme.

Confocal microscopy

Confocal images of the fluorescent samples were acquired via 20x (NA 0.75), 40x (NA 1.30) or 60x (NA 1.40) objectives of a Nikon A2 confocal microscope. The Z-stacked focal planes were taken and maximally projected by ImageJ, whereas brightness and contrast were adjusted in ImageJ if necessary.

Tile scans of whole coronal brain sections were captured with an Olympus VS120 or VS200 system and exported from OlyVIA Software in TIFF format. Further data analysis was carried out in ImageJ.

Two-photon surgery and imaging

Two-photon cranial window surgery was performed on mice anesthetized with isoflurane (3% induction, 1.2%-1.5% maintenance). A 3.5 mm-diameter circular cranial window was created at the skull above the sensory cortex, and a same-diameter circular D263T glass slice was implanted. The cranial window and a head bar were bonded using glue and dental cement to maintain stability. In addition to performing routine postoperative care, the mice were injected intraperitoneally with antibiotics and anti-inflammatory drugs (dexamethasone sodium phosphate injection and ceftiofur sodium for injection) for five consecutive days.

Two-photon imaging was performed on days 18, 22, 26, and 30 post-tamoxifen induction (as shown in the timeline in Fig. 2). During imaging, the mice were anesthetized with isoflurane (3% induction, 1.2%-1.5% maintenance), and body temperature was maintained by a heating pad. Imaging was performed via a self-developed two-photon microscopy system (developed by the Institute for Translational Brain Research, Fudan University) with one resonant and one galvo scan mirror. The excitation was delivered with a Coherent Chameleon Ti-Sapphire pulsed laser. The fluorescence emission was filtered with a Semrork FF01-520/35-25, FF01-593/46 filter and detected with a Hamamatsu photomultiplier tube (H7422-40P). The EGFP signal in Cx3cr1-CreER::LSL-Neurod1-EGFP mice was excited by a 920 nm laser and imaged via a 16x, 0.8 NA water-dipping objective (Nikon). Images were acquired at a depth of approximately 100 μm below the dura using ScanImage(40) at step intervals of 2 μm. Each Z-axis plane image (512x512 pixels) was imaged at a consistent zoom factor (∼300 × 300 μm area) and acquired at a 1.06 Hz frame rate.

Image processing was performed via ImageJ, and the Z-stack image containing the target cells was selected. Brightness and contrast were adjusted if needed, and the maximal projection was performed. Video processing was performed via Adobe Premiere software.

AAV vectors, lentivirus preparation, and stereotaxic injection

The AAV vectors were purchased from Obio Technology, Ltd., and PackGene Biotech, Inc., and were as follows: AAV1/2-CMV-EGFP, AAV5/2-CMV-EGFP, AAV6/2-CMV-EGFP, AAV8/2-CMV-EGFP, AAV9/2-CMV-EGFP, AAV2-DJ-CMV-EGFP, AAV2-retro-CMV-EGFP, AAV-PHP.eB-CMV-EGFP, AAV6-CX3CR1-GFP, AAV8-CX3CR1-GFP, AAV9-CX3CR1-GFP, AAV6-CD68-GFP, AAV8-CD68-GFP, and AAV9-CD68-GFP.

All lentiviruses were packaged by Obio Technology, Ltd., according to the author’s design and requirements. All lentiviruses were packaged in HEK293 cells, and the packaged plasmids consisted of pMDL, VSV-G and pREV. The following lentiviral vectors were used: pLenti-CAG-flag-NeuorD1, pLenti-CAG-flag-MCS, pLenti-CMV-GFP, pLenti-CX3CR1-GFP and pLenti-CD68-GFP.

For intracranial injection, the mouse was anesthetized with isoflurane. The stereotaxic injection coordinates were +1.0 mm anterior/posterior, ± 1.7 mm medial/lateral, and -0.75 mm dorsal/ventral.

Traumatic brain injury (TBI)

TBI was induced using the controlled cortical impact (CCI) model as previously described(36). Briefly, mice were anesthetized with ketamine (100 mg/kg) and xylazine (10 mg/kg) and placed in a stereotaxic frame (Stoelting, Wood Dale, IL, USA). Body temperature was maintained at 37.0 ± 0.5 °C using a heating pad throughout the procedure. After scalp disinfection, a midline incision (1.5–2 cm) was made to expose the skull. A 4 mm craniotomy was performed over the right parietal cortex (between bregma and lambda, 1 mm lateral to the midline) using a portable drill, taking care to preserve the dura mater. Mice with damaged dura were excluded from further analysis. Cortical impact was delivered using a 3 mm rounded steel impactor (PinPoint PCI3000, Hatteras Instruments Inc., USA) at a velocity of 1.5 m/s, impact depth of 1.5 mm, and dwell time of 100 ms. After impact, the bone flap was sealed with sterile bone wax and the scalp was sutured. Sham-operated mice underwent the same surgical procedures without cortical impact. Following surgery, mice were placed in a 37 °C recovery chamber until fully ambulatory.

Isolation of brain CD11b+ cells by magnetic-activated cell sorting (MACS)

Briefly, mice were deeply anesthetized with isoflurane and transcardially perfused with cold normal saline. Brains were harvested, rinsed briefly in cold normal saline, and mechanically minced using a blade in a brain mold. The minced tissue was transferred into a C-tube containing 3 ml of digestion buffer per brain. The digestion buffer consisted of DMEM (Gibco, REF: C11995500BT) supplemented with papain (8 U/ml, Worthington, cat: LS00312), DNase I (125 U/ml, Sangon, cat: B100649-0040), and trypsin inhibitor (1.5 mg/ml, SIGMA, cat: T9253 or Aladdin, cat: A274384). Tissue dissociation was performed using a gentleMACS Octo Dissociator with Heaters (Miltenyi, 37°C_ABCK1 program). The resulting cell suspension was gently triturated, and the C-tube was rinsed with 0.5% BSA in DMEM. The suspension was passed through a 70-μm cell strainer (Falcon, cat: CLS431751) and collected in a 50-ml tube. Cells were centrifuged at 300 × g for 5 min at 4°C, and the pellet was resuspended in 4 ml of 30% Percoll (Cytiva, cat:17544502) prepared in 0.5% BSA in DMEM. Following centrifugation at 300 × g for 5 min at 4°C, the supernatant was carefully removed, and the cell pellet was washed once with 500 μl of 0.5% BSA in DPBS by centrifugation at 300 × g for 5 min at 4°C. The cell pellet was then resuspended in 90 μl of 0.5% BSA in DPBS and incubated with 10 μl of anti-CD11b magnetic beads (Miltenyi, cat: 130-097-142) for 30 min at 4°C. The labeled cells were loaded onto an LS Column (Miltenyi, cat: 130-042-401) pre-equilibrated with 10 ml of 0.5% BSA in DPBS and washed with an additional 10 ml of the same buffer. The column was subsequently removed from the magnetic separator, and the magnetically labeled CD11b+ cells were eluted with 5 ml of 0.5% BSA in DPBS by firmly applying the plunger twice. The eluted cells were collected and centrifuged at 300 × g for 5 min at 4°C, followed by resuspension in the appropriate buffer for cell counting and downstream experiments.

Rotarod test

All mice were acclimated to the behavioral room for half an hour before test. Briefly, in the training phase, the mice were placed on rotating rod at 5 revolution per minute (RPM) to learn how to walk forward and maintain balance. Each mouse was placed on the rod for 60 seconds and then returned back to its home cage. This procedure was repeated three times with 10 minuties intervals. In the testing phase, mice were placed in separate lanes on the rod, which was set to accelerate from 4 to 40 RPM over a 300-second. This test was repeated three times, with approximately 30 minutes intervals between trails. The latency to fall was recorded and used for subsequent statistical analysis.

Grip strength test

All mice were acclimate to the behavioral room for half an hour before test. In brief, we gently held mouse by the tail and placed it with all four paws on the grip strength grid. The mouse was then gently pulled back until it relased the grasp, and the holding force was recorded. This mearsurement was repeated six times per mouse, and the maximum force value was used for subsequent statistical analysis.

Single-cell preparation and fluorescence-activated cell sorting (FACS)

For scRNA-seq, FACS was used to harvest GFP+ or tdTomato+ cells from tamoxifen-treated Cx3cr1-creER::LSL-Neurod1-GFP and Cx3cr1-CreER::Ai14 mice. Briefly, brains were dissected and cut into 1-mm³ pieces using a mouse stainless steel brain matrix (RWD). The tissue pieces were digested in 3 ml of DMEM containing 8 U/ml papain (8 U/ml, Worthington, cat: LS00312), 125 U/ml DNase I (Sangon Biotech, cat: B100649-0040), and 1.5 mg/ml trypsin inhibitor (SIGMA, cat: T9253 or Aladdin, cat: A274384) via a gentleMACS Octo Dissociator (Miltenyi) with the 37C_ABDK program.

After 30 min of digestion, the mixture was gently pipetted up and down and then filtered through a 70 µm cell strainer. The dissociated cells were centrifuged at 370 × g at RT for 10 min, and the supernatant was discarded. The cell pellets were resuspended in 4 ml of 30% Percoll (Cytiva, cat: 17089101) and centrifuged again at 370 × g and 17°C for 10 min to remove debris. After centrifugation, the supernatant was discarded, and the cells were washed once with DPBS containing 0.5% BSA. Before the cells were loaded onto the SONY MA900, they were stained with 7-AAD (BD Pharmingen, cat: 559925, 1:700) to label dead cells. Finally, the tdTomato+ 7-AAD- or GFP+ 7-AAD- cells were collected for scRNA-seq immediately.

scRNA-seq data analysis

scRNA-seq was conducted via the MobiCube High-throughput Single Cell 3 ′ Transcriptome Set V2.1 (PN-S050200301) and the MobiNova-100 microfluidic platform. The single-cell suspension was adjusted to a concentration of 700–1200 cells/μl and immediately loaded onto a chip for microdroplet formation via the MobiNova-100 system. Reverse transcription, cDNA amplification, and DNA library construction were carried out according to the manufacturer’s protocol. High-throughput sequencing was performed in PE-150 mode.

FASTQ files were processed and aligned to the mouse reference genome (mm10) via MobiVision version 3.2, with unique molecular identifier (UMI) counts aggregated for each barcode. The resulting UMI count matrix was analyzed in R version 4.2.2 via Seurat version 4.3.0. To ensure data quality, we excluded low-quality cells and potential multiplets on the basis of the following criteria: cells with fewer than 200 genes or more than 5000 genes and cells with more than 10% UMIs mapped to mitochondrial genes. scRNA-seq data from the four groups were integrated via IntegratedData functions according to the manufacturer’s instructions. The remaining data were normalized via the NormalizeData function and scaled with ScaleData. For scRNA-seq analysis, principal component analysis (PCA) was conducted on variable features, with the top 30 principal components (PCs) selected for clustering and visualization via tSNE and uniform manifold approximation and projection (UMAP). Cell clustering was performed with a resolution of 1 to annotate the cell types. Differentially expressed genes (DEGs) among cell groups were identified via the FindMarkers function with parameters set to min.pct = 0.1 and a log fold change (FC) threshold of 0.25 (Wilcoxon rank-sum test). Genes with |log2FC| > 0.25 and P < 0.05 were considered significant DEGs and subsequently used for Gene Ontology (GO) analysis. Visualization of the results was performed via the VlnPlot and FeaturePlot functions. We gratefully thank OE Biotech Co., Ltd. (Shanghai, China) for providing sequencing service.

Statistical analysis

Statistical analyses were performed with Prism 10.2.0 (GraphPad). Each data point represents the average statistical result of more than three brain sections in different bregma regions of one mouse. The results were evaluated independently in a double-blind manner. All the results are presented as the means ± standard deviations (SDs). One-way analysis of variance (ANOVA) with Holm‒Sidak’s multiple comparisons test (post hoc) was performed for multiple comparisons, whereas a two-tailed independent t test was used to compare the differences between two groups. No outliers were excluded.